---
title: "chngpt means ReadMe"
author: "Bart Huntley"
date: "Monday, May 25, 2015"
output: html_document
---

This function uses the "*changepoint*" package to define breaks in means of a time series trace. Various methods can be called within the cpt.mean call. For further details on this call and the changepoint package refer to <https://www.google.com.au/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=0CB0QFjAA&url=http%3A%2F%2Fcran.r-project.org%2Fweb%2Fpackages%2Fchangepoint%2Fchangepoint.pdf&ei=V5diVf_4E8WMmwWmlIH4Aw&usg=AFQjCNGkR3zx2bsbbYA4JfSRtExNhFAikQ&sig2=tnWVwxUu8wRS37Lnix7Ftg>.

### Notes

The starting point for this function is a **...mtsd.csv** file. This .csv file is the result of running **mtsd.r** on cloud QA'd data - see **mtsd_ReadMe** for further information.

The output ggplots will be saved to a folder named "ts-chgpt-graphs-TODAYS DATE". Running this function on another day will save the output to a folder with that days date. This is an attempt to help keep iterations of graphs clearly identifiable.

The output ggplot dimensions have been optimised for the creation of .pdf ggplots that fit a standard word doc in portrait layout. If you are creating outputs with different file extensions (see step 3 for details) or different page layouts and the dimensions do not suit, these can be adjusted in the final line of the function (line 109). 

Please do not save changes to this script. If you need to tweak the script for specific project requirements, please save a local copy of this script in the project folder. This will aid repeatability of the analysis and maintain the integrity of the original script.



### Details
1. Step 1 - Run the code to create the function
```{r, eval=FALSE}
chgpt_means <- function(dir, csv, survey, out){
        
        is_installed <- function(mypkg) is.element(mypkg, installed.packages()[,1])
        load_or_install<-function(package_names)  
        {  
                for(package_name in package_names)  
                {  
                        if(!is_installed(package_name))  
                        {  
                                install.packages(package_name,repos="http://cran.csiro.au/")  
                        }  
                        library(package_name,character.only=TRUE,quietly=TRUE,verbose=FALSE)  
                }  
        }  
        load_or_install(c("lubridate","ggplot2", "dplyr","tidyr", "grid", "gridExtra",
                          "changepoint"))
        
        ##Generic tasks
        setwd(dir)
        df <- read.csv(csv, header = TRUE)
        df <- df[,-1]
        df[,1] <- as.Date(df[,1])
        
        sname <- names(df)[-1]
        folder <- paste0("ts-chgpt-graphs-", Sys.Date())
        if(!file.exists(folder)){ dir.create(folder)}
        dir.create(folder)
        setwd(paste(dir,folder, sep="\\"))

        for (i in 1:length(sname)){
                
                df2.i <- df[, c(1, 1+i)]
                df2.i$label <- factor(rep("ts", by=length(df2.i[,1])))
                
                #Helper function
                indexer <- function(){
                        ev <- vector(mode="numeric", length=length(cpts))
                        for(j in 1:length(cpts)){
                                ev[j] <- as.numeric(df2.i[cpts[j],1])
                        }
                        ev
                }
                #Changepoint
                site.ts <- ts(df2.i[,2], frequency=12, 
                              start=c(as.numeric(year(df2.i[1,1])),
                                      as.numeric(month(df2.i[1,1]))),
                              end=c(as.numeric(year(tail(df2.i[,1], n=1))),
                                    as.numeric(month(tail(df2.i[,1], n=1)))))
                mvalue <- cpt.mean(site.ts, method="BinSeg")
                
                cpts <- mvalue@cpts
                
                #raw dates of chgpts - mix of star and end dates for periods
                date <- c(df[1,1], as.Date(indexer()))
                
                cptdf <- data.frame(start=date[1:length(date)-1],
                                    end=date[-1], y=mvalue@param.est$mean, label=factor("mean"))
                #df for vertical line handling
                vertdf <- data.frame(x=as.Date(survey), y=c(-Inf, Inf),
                                     Survey=factor(year(as.Date(survey))))
                
                p <- ggplot()+
                        geom_point(data=df2.i, aes_string(x="date", y=sname[i], colour="label"))+
                        geom_line(data=df2.i, aes_string(x="date", y=sname[i], colour="label"))+
                        scale_colour_manual(values=c("black"),
                                            name=sname[i],
                                            breaks=as.character(df2.i$label),
                                            labels=as.character(df2.i$label))+
                        geom_segment(data=cptdf, aes(x=start, y=y, xend=end, yend=y), colour="red")+
                        geom_line(aes(x,y, linetype=Survey), colour='blue', size = 0.5, vertdf)+
                        coord_cartesian(ylim = c(-10, 80))+
                        theme_bw()+
                        xlab("")+
                        ylab("Vegetation Cover %")+
                        theme(legend.justification=c(0,1), 
                              legend.position=c(0,1),
                              axis.title.y= element_text(size=15),
                              axis.text.y = element_text(angle=90, size=15),
                              axis.text.x = element_text(size=15),
                              legend.title = element_text(size = 15),
                              legend.text = element_text(size = 15)) 
                
                
                sname.i<-sname[i]
                filename<-paste0(sname.i, "-chgpt-plot",out)

                ggsave(file=filename, p, width = 22.5, height = 13.5, units = "cm")
        }
}
```

2. Step 2 - Define the function arguments (see examples)
    + dir = working directory filepath where function script and **...mtsd.csv** reside. Ensure backslashes are doubled and filepath is encapsulated in " "
    + csv = name of .csv file, including .csv extension, encapsulated in " "
    + out = output file extension. Suggest ".pdf" for scalable vector graphics however can be ".jpeg"", ".png"" etc.
    + survey = "yyyy-mm-dd" of a date of significance, in this case when a veg survey was conducted. Input is used to place a vertical line on the output plot.
```{r, eval=FALSE}
dir="Z:\\DEC\\Vegmachine_SharkBay\\Working\\Bernier_Dorre\\Working\\analysis"
csv="Bernier_Dorre_mtsd.csv"
out=".pdf"
survey="1998-09-01"
```

3. Step 3 - Run the function call
```{r, eval=FALSE}
chgpt_means(dir, csv, out, survey)
```

###Errors
If an error appears try
```{r, eval=FALSE}
rm(list=ls())
```
to clear the working environment and re-run the steps in order. If problems persist have a chat to Bart.


